Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-09 paper_jasa.tex
changed axis label to use greek letter, size of figure 3
lemma 3.1

The probability of picking the data plot from a lineup depends on the size $m$ of the lineup and the strength of the signal in the data plot. 
Under the above assumptions, we can express the probability as:
\[
P(p_D < p_0) = \left(1 - E[p_D] \right)^{m-1} 
\]
where $p_D$ is the $p$-value associated with the data in the test statistic, and $p_0$ is the minimum of all $p$-values in the data going into null plots.
\end{lemma}

@@

Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
\begin{eqnarray*}
P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
\end{eqnarray*}
which, in particular, is independent of $p_{0,i}$ for all $i$.

Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.

With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.

Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
\[
P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
\]
\end{proof}

@@

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% Equation \ref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \ref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.65}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of UMP test for sample size $n =100$, $\sigma = 12$ and $m$=20. \green{Produce these curves for $m=12$, $m=20$, $m=30$.}}
       \label{fig:power_expected}
\end{figure}

>>>>

The probability of picking the data plot from a lineup depends on the size $m$ of the lineup and the strength of the signal in the data plot. 
Under the above assumptions, we can express the probability as:
\[
P(p_D < p_0) = 1 - E\left[ (1 - p_D)^{m-1}\right]
\]
where $p_D$ is the $p$-value associated with the data in the test statistic, and $p_0$ is the minimum of all $p$-values in the data going into null plots.
\end{lemma}

@@

\hh{Their minimum $p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}$ then follows a Beta distribution with shape parameters 1 and $m-1$, and corresponding distribution function 
\[
F_{p_0} (x) =  (1- x) ^{m-1} \text{ for } x \in [0,1].
\]

\begin{eqnarray*}
P(p_D < p_{0}) &=& 1 - P(p_{0} \le p_D) = 1- \int_0^1  P(p_{0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
&=& 1 - \int_0^1 F_{p_{0}}(t) f_{p_D}(t) dt = 1 - \int_0^1 (1-t)^{m-1} f_{p_D}(t) dt = 1 - E\left[ (1 - p_D)^{m-1}\right],
\end{eqnarray*}

}
%
%
%
%Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
%We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
%\begin{eqnarray*}
%P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
%&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
%\end{eqnarray*}
%which, in particular, is independent of $p_{0,i}$ for all $i$.
%
%Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.
%
%With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.  
%
%Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
%\[
%P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
%\]
\end{proof}

@@

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% Equation \ref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \ref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.95}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of UMP test for sample size $n =100$, $\sigma = 12$ and $m$=20. }
       \label{fig:power_expected}
\end{figure}
